\documentclass[a4paper, twoside]{report}
\usepackage{subfiles}
\input{../preamble.tex}
\title{Dr. Widom's Quantum Mechanics I Lectures}
\date{Fall Semester, 2019}
\begin{document}
    \maketitle
    \tableofcontents
    \chapter*{Preface}
    These are my notes from the lecture series taught by Dr. Michael Widom at Carnegie Mellon University during the Fall of 2019. I hope these notes will prove useful to current and future students of the course. Please note that I am missing lecture $ 20 $, which I believe concerned quantum computing. Supplemental notes for this lecture were included online by Dr. Widom, so I did not feel the need to retype them here. I cannot guarantee the accuracy of these notes, as they were all typed (quickly) during class with very minimal proofreading afterwards. Additionally there may be some missing information towards the end of the first twelve lectures (which were originally typed in a different format and converted to \LaTeX\ later). I have tried to organize the notes roughly into chapters, and these chapters do not reflect the sections in \textit{Consistent Quantum Theory} (Robert B. Griffiths) or \textit{Quantum Mechanics, Volume 1} (Claude Cohen-Tannoudji et al.), the textbooks used in this course.

        - Dene Hoffman

    \subfile{lec_1-12.tex}
    \subfile{lec_13.tex}
    \subfile{lec_14.tex}
    \subfile{lec_15.tex}
    \subfile{lec_16.tex}
    \subfile{lec_17.tex}
    \subfile{lec_18.tex}
    \subfile{lec_19.tex}

    \subfile{lec_21.tex}
    \subfile{lec_22.tex}
    \subfile{lec_23.tex}
    \subfile{lec_24.tex}
    \subfile{lec_25.tex}
    \subfile{lec_26.tex}
    \subfile{lec_27.tex}
    \subfile{lec_28.tex}
    \subfile{lec_29.tex}
    \subfile{lec_30.tex}
    \subfile{lec_31.tex}
    \subfile{lec_32.tex}
    \subfile{lec_33.tex}
    \subfile{lec_34.tex}
    \subfile{lec_35.tex}
    \subfile{lec_36.tex}
    \subfile{lec_37.tex}
    \subfile{lec_38.tex}
    \subfile{lec_39.tex}
    \subfile{lec_40.tex}
    \subfile{lec_41.tex}
    \subfile{lec_42.tex}
    \subfile{lec_43.tex}
    \subfile{lec_44.tex}
    \subfile{lec_45.tex}

\end{document}
